Symbolic Abstract Data Type Inference

### 2015-07-08 draft.tex
Convincing intro.

  Formal specifications are a vital component to the scalable verification of
  software systems. In the case of efficient implementations of concurrent
  objects like atomic registers, queues, and locks, symbolic representations
  of their abstract data types (ADTs) enable efficient modular reasoning,
  decoupling clients from implementations. Writing adequate formal
  specifications, however, is a complex task requiring rare expertise. In
  practice, concurrent objects are specified indirectly by reference
  implementations.

>>>>

  Formal specification is a vital ingredient to scalable verification of
  software systems. In the case of efficient implementations of concurrent
  objects like atomic registers, queues, and locks, symbolic formal
  representations of their abstract data types (ADTs) enable efficient modular
  reasoning, decoupling clients from implementations. Writing adequate formal
  specifications, however, is a complex task requiring rare expertise. In
  practice, programmers write reference implementations as informal
  specifications.

### 2015-07-08 intro.tex
Convincing intro.

Many modern software systems rely on efficient implementations of abstract data
types (ADTs) like atomic collections, semaphores, and locks. Efficient
implementation is difficult though, because efficiency entails maximizing
concurrency, and concurrency jeopardizes atomicity. Consequently,
implementations are vulnerable to subtle and elusive bugs compromising
adherence to their ADTs~\cite{journals/cacm/staff11}, thus necessitating
automated verification. Adherence of an implementation to an ADT is captured
formally by a property called \emph{observational refinement}: any observable
behavior over the method call and return actions of the implementation is also
admitted by the the ADT.

Although checking observational refinement is a fundamentally hard
problem\footnote{See Section~\ref{sec:related} for further discussion.} for
arbitrary ADTs~\cite{journals/siamcomp/GibbonsK97, conf/esop/BouajjaniEEH13},
recent work demonstrates that implementations of naturally-occurring ADTs like
atomic collections, semaphores, and locks can be checked efficiently given
effective symbolic ADT representations~\cite{conf/popl/BouajjaniEEH15,
journals/arxiv/BouajjaniEEH15, conf/pldi/EmmiEH15}, e.g.,~specified in
first-order logic. However, writing accurate symbolic specifications is an
expected challenge for ADT designers. In practice, ADTs are specified
indirectly as the observable behaviors of possibly-abstract imperative
\emph{reference implementations}. Mapping these reference implementations to
logical formulæ describing the same sets of behaviors is a complex task
requiring rare expertise.

In this work we demonstrate that effective symbolic ADT representations can be
generated automatically from the executions of reference implementations,
enabling the application of efficient symbolic refinement-checking algorithms
without the burden of writing formal specifications manually. Our approach
exploits two key features of concurrent-object ADTs: that violations of each
ADT can be decomposed into a small set of representative patterns, and that
these patterns manifest in executions with few operations. The first feature
allows us to represent symbolic ADTs finitely, as exclusions of violation
patterns. The second allows us to extract violation patterns from finite
enumerations of executions.

@@

Computing the basis sets of ADT violation patterns is a challenging problem,
requiring the computation of global properties of an infinite number of
executions — analogously to the inference of inductive invariants. Exploiting a
hypothesis that violation patterns manifest in executions with few operations,
we propose an under-approximating algorithm which extracts the patterns
observed in all violating executions up to a given number of operations. In
theory, for an arbitrary ADT, this is clearly incomplete: any violation which
only surfaces with a greater number of operations would not be captured, thus
resulting in a symbolic ADT representation which can fail to identify
violations — though still guarantees never to classify a valid execution as a
violation. Empirically, however, we demonstrate that our hypothesis holds: the
patterns emerging from executions with few operations are complete in
characterizing all violations of naturally-occurring concurrent object ADTs,
thus allowing us to compute complete symbolic ADT representations in practice.

Although our approach does require annotating the operations of concurrent
object executions with a matching relation, and we demonstrate that these
relations can be given for naturally-occurring ADTs, we also provide an
automatic means for computing such \emph{matching schemes}. Again by sampling
executions, we leverage automated symbolic reasoning engines to synthesize
matching schemes for which given implementations are closed under the removal
of matches. This further lowers the burden of automated verification. Rather
than providing formal ADT specifications, or even matching schemes, users need
only provide the predicates relevant in the logic of matching schemes, and we
could automatically compute effective matching schemes, and ultimately,
effective symbolic ADT representations.

@@

  \item The statement of the ADT inference problem (§\ref{sec:inference}).

  \item Identification of the algebraic properties allowing finite
  characterization and computability of infinite ADT violations sets
  (§\ref{sec:patterns}).

  \item The computability of symbolic ADT representations (§\ref{sec:formula}).

@@

To the best of our knowledge, this work is the first to suggest the automatic
generation of symbolic ADT representations for concurrent objects. By removing
the burden of writing formal ADT specifications manually, this work broadens
the scope of efficient symbolic refinement-checking algorithms to
newly-designed ADTs, apart from those few traditionally studied in the
literature.

>>>>

Effective scalable reasoning about nontrivial software implementations generally
requires considering each software module separately, in isolation, using
abstract specifications for other modules. When modules are objects whose
methods may be called concurrently, their behavior is typically understood in
terms of invocation sequences of abstract data types (ADTs): an execution with
overlapping method invocations is considered valid when those invocations can be
\emph{linearized} into a sequence admitted by the
ADT~\cite{journals/toplas/HerlihyW90}. For example, consider the execution
history depicted in Figure~\ref{fig:clients} in which the add operations
numbered $2$ and $3$ overlap with each other, and, respectively, with operations
$1$ and $4$. This execution is valid with respect to the atomic queue ADT
because among the five possible ways to linearize its six operations, the
sequence~$1$, $3$, $2$, $4$, $5$, $6$ is admitted. ADT specifications thus
decouple reasoning about object implementations from their clients’ invocations:
\begin{itemize}

  \item Does every implementation execution have a valid linearization?

  \item Does every valid linearizations preserve client invariants?

\end{itemize}
The former question depends only on a given object’s implementation, and the
latter only on a given object’s clients.

\begin{figure}[t]
  \begin{minipage}{0.43\linewidth}
    \begin{verbatim}
1: add(a)       #
2: add(b)       ###
3: add(c)         ###
4: add(d)           #
5: remove => a        #
6: remove => c          #
    \end{verbatim}
  \end{minipage}
  \hfill
  \begin{minipage}{0.55\linewidth}
    \begin{verbatim}
assume w < x
(  (q.add(<1,w>); q.add(<1,x>))
|| (q.add(<2,w>); q.add(<2,x>)) )
i, y := q.remove()
j, z := q.remove()
assert i == j ==> y < z
    \end{verbatim}
  \end{minipage}
  \caption{An execution history with six numbered operations (left),
    and a parallel program invoking six operations (right). The “\#” symbols
    depict the time intervals spanned by operations horizontally.}
  \label{fig:clients}
\end{figure}

Although formal ADT specifications are indispensable for scalable program
reasoning, formal-specification writing is a burden for which few programmers
posses the required combination of expertise and willingness to overcome.
Typically programmers write simple ADT \emph{reference implementations},
e.g.,~whose methods are synchronized via a global lock, and refine them with
more efficient fine-grained implementations, e.g.,~reducing synchronization
bottlenecks using specialized hardware instructions such as atomic compare and
swap (CAS).

Our goal in this work is thus the automated generation of formal ADT
specifications, derived from reference implementations, which are suitable for
automated reasoning. In particular, we aim to generate \emph{symbolic
representations} of valid invocation sequences for ADTs which are given
implicitly by reference implementations. We target declarative symbolic
representations rather than imperative state-based representations in order to
harness efficient symbolic reasoning algorithms: rather than enumerating
linearizations explicitly, and checking their validity one by one, a symbolic
reasoning engine may simultaneously rule out many possible linearizations.
Previous work demonstrates that such symbolic reasoning can increase
efficiency by orders of magnitude~\cite{conf/pldi/EmmiEH15}.

In this work we demonstrate that effective symbolic ADT representations can be
generated automatically from the executions of reference implementations,
enabling the application of efficient symbolic reasoning algorithms
without the burden of writing formal specifications manually. Our approach
exploits two key features of concurrent-object ADTs: that violations of each
ADT can be decomposed into a small set of representative patterns, and that
these patterns manifest in executions with few operations. The first feature
allows us to represent symbolic ADTs finitely, as exclusions of violation
patterns. The second allows us to extract violation patterns from finite
enumerations of executions.

@@

Computing the basis sets of ADT violation patterns is a challenging problem,
requiring the computation of global properties of an infinite number of
executions — analogously to the inference of inductive invariants. Exploiting a
hypothesis that violation patterns manifest in executions with few operations,
we propose an under-approximating algorithm which extracts the patterns observed
in all violating executions up to a given number of operations. In theory, for
an arbitrary ADT, this is clearly incomplete: any violation which only surfaces
with a greater number of operations would not be captured, thus resulting in a
symbolic ADT representation which can fail to identify violations — though still
guarantees never to classify a valid execution as a violation, and is thus sound
for program reasoning. Empirically, however, we demonstrate that our hypothesis
holds: the patterns emerging from executions with few operations are complete in
characterizing all violations of naturally-occurring concurrent object ADTs,
thus allowing us to compute complete symbolic ADT representations in practice.

Although our approach does require annotating the operations of concurrent
object executions with a matching relation, and we demonstrate that these
relations are easy to provide for naturally-occurring ADTs, we also outline an
automatic means for computing such \emph{matching schemes}. Again by sampling
executions, we leverage automated symbolic reasoning engines to synthesize
matching schemes for which given implementations are closed under the removal
of matches. This further lowers the burden of automated verification. Rather
than providing formal ADT specifications, or even matching schemes, users need
only provide the predicates relevant in the logic of matching schemes, and we
could automatically compute effective matching schemes, and ultimately,
effective symbolic ADT representations.

@@

  \item The statement of the symbolic ADT inference problem (§\ref{sec:inference}).

  \item Identification of the algebraic properties allowing a finite
  characterization of infinite ADT violation sets (§\ref{sec:patterns}).

  \item The symbolic representation of ADT violations (§\ref{sec:formula}).

  \item The computability of symbolic ADT representations (§\ref{sec:algorithm}).

@@

To the best of our knowledge, this work is the first to suggest the automatic
generation of ADT specifications. By removing the burden of writing formal ADT
specifications manually, this work broadens the scope of modular program
reasoning using efficient symbolic algorithms to newly-designed ADTs, apart from
those few traditionally studied in the literature.

### 2015-07-08 motivation.tex
Comment on the possible queue violations.
Convincing intro.

Effective scalable reasoning about nontrivial software implementations generally
requires considering each software module separately, in isolation, using
abstract specifications for other modules. When modules are objects whose
methods may be called concurrently, their behavior is typically understood in
terms of valid invocation sequences of atomic abstract data types (ADTs): any
execution of overlapping method invocations can be \emph{linearized} into an
invocation sequence admitted by the ADT~\cite{journals/toplas/HerlihyW90}. For
example, consider the execution history depicted in Figure~\ref{fig:clients} in
which the add operations numbered $2$ and $3$ overlap with each other, and,
respectively, with operations $1$ and $4$. Among the five possible ways to
linearize these six operations, the linearization~$1$; $3$; $2$; $4$; $5$; $6$
does correspond to a valid sequence of the atomic queue ADT, and so this history
is linearizable. ADT specifications thus decouple reasoning about object
implementations from their client’s invocations:
\begin{itemize}

  \item Does every implementation execution have a valid linearization?

  \item Does every valid linearizations preserve client invariants?

\end{itemize}
The former question depends only on a given object’s implementation, and the
latter only on a given object’s clients.

\begin{figure}[t]
  \begin{minipage}{0.43\linewidth}
    \begin{verbatim}
1: add(a)       #
2: add(b)       ###
3: add(c)         ###
4: add(d)           #
5: remove => a        #
6: remove => c          #
    \end{verbatim}
  \end{minipage}
  \hfill
  \begin{minipage}{0.55\linewidth}
    \begin{verbatim}
assume w < x
(  (q.add(<1,w>); q.add(<1,x>))
|| (q.add(<2,w>); q.add(<2,x>)) )
i, y := q.remove()
j, z := q.remove()
assert i == j ==> y < z
    \end{verbatim}
  \end{minipage}
  \caption{An execution history with six numbered operations (left),
    and a parallel program invoking six operations (right). The “\#” symbols
    depict the time intervals spanned by operations horizontally.}
  \label{fig:clients}
\end{figure}

@@

Although formal ADT specifications are indispensable for scalable program
reasoning, formal-specification writing is a burden for which few programmers
posses the required combination of expertise and willingness to overcome.
Typically programmers write simple ADT \emph{reference implementations},
e.g.,~whose methods are synchronized via a global lock, and refine them with
more efficient fine-grained implementations, e.g.,~reducing synchronization
bottlenecks using specialized hardware instructions such as atomic compare and
swap (CAS).

Our goal in this work is thus the automated generation of formal ADT
specifications, derived from reference implementations, which are suitable for
automated reasoning. In particular, we aim to generate \emph{symbolic
representations} of the valid history linearizations, according to ADTs given
implicitly by reference implementations. We target declarative symbolic
representations rather than imperative state-based representations in order to
harness efficient symbolic reasoning algorithms: rather than enumerating
linearizations explicitly, and checking their validity one by one, a symbolic
reasoning engine may simultaneously rule out many possible linearizations,
e.g.,~as in the example above. Previous work demonstrates that such symbolic
reasoning can increase reasoning efficiency by orders of
magnitude~\cite{conf/pldi/EmmiEH15}.

@@

As an example of this algorithm at work, consider the sequential histories
listed in Figure~\ref{fig:patterns}. These sequences arise from an enumeration
of all $202$ possible method invocation sequences\footnote{Here we consider
equivalence among sequences which are isomorphic up to renaming of data values,
e.g.,~to avoid considering {\tt add(1); add(2); remove => 1} and {\tt add(3);
add(2); remove => 3} as distinct. The notion of \emph{matching} introduced in
Section~\ref{sec:histories} provides a clean formal treatment.} of length at
most $4$ of an object with add and remove methods for which remove can return
empty. The $31$ sequences which are executable by a correct reference
implementation of an atomic queue implementation --- i.e.,~with matching return
values for all invocations --- are discarded. The remaining $171$ sequences are
invalid, $164$ of which are redundant with the seven {\tt PATTERNS} listed on the
left-hand side of Figure~\ref{fig:patterns}. Seven of these redundant sequences
are shown on the right-hand side of Figure~\ref{fig:patterns}. For example, the
first five sequences on the right, as well as the last, are redundant with the
first on the left since they each describe a violation in which a removed
element was never added. The sixth sequence on the right is redundant with the
second on the left since they both describe a violation in which remove returns
empty before previously-added elements are removed. Finally, the exclusion of
the computed pattern set can be expressed as a conjunction of formulas
in first order logic describing the exclusion of each individual pattern. For
example, the first pattern could be described by the formula
\begin{align*}
  \exists o.\
    {\sf method}(o) = {\tt remove} \land {\sf unmatched}(o)
\end{align*}
describing an unmatched remove operation.

>>>>

@@

TODO SEGUE

@@

As an example of this algorithm at work, consider the sequential histories
listed in Figure~\ref{fig:patterns}. These sequences arise from an enumeration
of all $202$ possible method invocation sequences\footnote{Here we consider
equivalence among sequences which are isomorphic up to renaming of data values,
e.g.,~to avoid considering {\tt add(1); add(2); remove => 1} and {\tt add(3);
add(2); remove => 3} as distinct. The notion of \emph{matching} introduced in
Section~\ref{sec:histories} provides a clean formal treatment.} of length at
most $4$ of an object with add and remove methods for which remove can return
empty. The $31$ sequences which are executable by a correct reference
implementation of an atomic queue implementation --- i.e.,~with matching return
values for all invocations --- are discarded. The remaining $171$ sequences are
invalid, $164$ of which are redundant with the seven {\tt PATTERNS} listed on
the left-hand side of Figure~\ref{fig:patterns}. Seven of these redundant
sequences are shown on the right-hand side of Figure~\ref{fig:patterns}. For
example, the first five sequences on the right, as well as the last, are
redundant with the first on the left since they each describe a violation in
which a removed element was never added. The sixth sequence on the right is
redundant with the second on the left since they both describe a violation in
which remove returns empty before previously-added elements are removed. Other
represented violations include removing an element before it is added, and
removing elements in the opposite order in which they were added. Finally, the
exclusion of the computed pattern set can be expressed as a conjunction of
formulas in first order logic describing the exclusion of each individual
pattern. For example, the first pattern could be described by the formula
\begin{align*}
  \exists o.\
    {\sf method}(o) = {\tt remove} \land {\sf unmatched}(o)
\end{align*}
describing an unmatched remove operation.
